Computational methods in landscape ecology

### 2024-09-19 Sections/03-connectivity.qmd
Minor cosmetic changes

Landscape connectivity describes how landscape features facilitate or impede movement [@Taylor2006LandscapeConnectivity].
It underpins ecosystem functionality, maintains biodiversity and populations, and plays an important role in many conservation actions [@CorreaAyram2016HabitatConnectivity].
Connectivity can be divided into structural and functional connectivity.
Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency.
It is solely a landscape feature and independent of species' characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioral, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].

Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, quantifying connectivity has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
Movement data from many individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; @Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
Thus, in practice connectivity is primarily determined by indirect estimations.
@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.

## Estimating structural connectivity

Landscape metrics are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, however, functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].

@@

A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapejlSoftware], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].

Least-cost paths assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a threshold.
The creation of least-cost paths is a well established technique and there are highly optimized and efficient algorithms available for its calculation.

@@

However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

@@

Following the formative work of @Urban2001LandscapeConnectivity, graph theory has developed into a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement in landscape graphs [@Cushman2013BiologicalCorridors].
Graph theory and connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
For example, it can assist wildlife movement corridor planning [e.g., @Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity for protection and restoration [@Liu2020UsingLandscape].

Within the landscape delineating the system boundary, discrete habitat patches are modeled as nodes, and the potential movement between patches is modeled as edges.
The edges can be binary, representing whether a connection between two nodes exists or not, or weighted by cost values representing movement efforts from one patch to another [@Petsas2021LandscapeConnectivity].
Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.
Widely reported connectivity metrics include the probability of connectivity index [@Saura2007NewHabitat], habitat availability metrics quantifying potential and functional connectivity [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], or the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment] which is based on a binary connection model [@Saura2010CommonCurrency].

Recently, a new generation of graph-theoretical metrics, commonly reported in combination and addressing caveats of previous metrics [@Hashemi2022ReviewEcological], have been developed to account for landscape features critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
The modeling framework, initially only applied to single species is increasingly being used to model multispecies landscapes [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach], which account for interspecific movement abilities and habitat preferences.
In addition, modern multiple-layer graph approach offer to model not only spatial but spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
Modeling the temporal dynamics helps to identify habitat patches impact connectivity over time, and how these impacts affect biodiversity patterns [@Uroy2021HowAssess].

However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge due to various ways to construct a graph and different data sources.
In many cases, data availability limits the representation of the modelled landscape, with subsequent implications for the calculated connectivity metrics and inferences of connectivity [Foltete2020CouplingLandscape].
For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types resulting in different distribution of connectivity values.

Currently, graph-theoretical landscape graphs and connectivity indices can be calculated using several open-source software, including *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], the R packages *lconnect* [@Mestre2023LconnectPackage] and *grainscape* [@Chubaty2020ToolboxGrainscape], or Julia’s *ConScape* library [@vanMoorter2023AcceleratingAdvances], among others.

>>>>

Landscape connectivity describes how landscape features facilitate or impede movement [@Taylor2006LandscapeConnectivity] and can be divided into structural and functional connectivity.
It underpins ecosystem functionality, maintains biodiversity and populations, and plays an important role in many conservation actions [@CorreaAyram2016HabitatConnectivity].
Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency.
It is solely a landscape feature and independent of species characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioral, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].

Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, quantifying connectivity has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
Generally, movement data from individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; @Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
Thus, in practice connectivity is primarily determined by indirect estimations.
@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal characteristics of species, and iii) actual connectivity related to observations of individuals moving through a landscape.

## Structural connectivity

Landscape metrics are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, yet functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].

@@

Least-cost paths assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping to identify reachable areas within a threshold.
The creation of least-cost paths is a well established technique and there are highly optimized and efficient algorithms available for its calculation.

@@

However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

@@

Following the formative work of @Urban2001LandscapeConnectivity, graph theory has developed into a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement in landscape graphs [@Cushman2013BiologicalCorridors].
Graph theory and connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
For example, they can assist wildlife movement corridor planning [e.g., @Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].

Within the landscape, discrete habitat patches are modeled as nodes, and the potential movement between patches is modeled as edges.
The edges can be binary, representing whether a connection between two nodes exists or not, or weighted by cost values representing movement efforts from one patch to another [@Petsas2021LandscapeConnectivity].
Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.
Widely reported connectivity metrics include the probability of connectivity index [@Saura2007NewHabitat], habitat availability metrics quantifying potential and functional connectivity [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], or the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment] which is based on a binary connection model [@Saura2010CommonCurrency].

Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
Furthermore, the modelling framework initially only applied to single species is increasingly being used to model multispecies landscapes to account for interspecific movement abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
In addition, modern multiple-layer graph approaches offer to model not only spatial, but spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and how these impacts affect biodiversity patterns [@Uroy2021HowAssess].

However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge due to various ways how to construct a graph and different data sources.
In many cases, data availability limits the representation of the modelled landscape with subsequent implications for the calculated connectivity metrics and inferences of connectivity [Foltete2020CouplingLandscape].
For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types which resulted in different distribution of connectivity values.

A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [Julia\; @vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [Julia\;@Landau2021OmniscapejlSoftware], *lconnect* [R\; @Mestre2023LconnectPackage], *grainscape* [R\; @Chubaty2020ToolboxGrainscape], *Circuitscape* [@Anantharaman2019CircuitscapeJulia], *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], among others.

### 2024-09-19 Sections/05-simulations.qmd
Remove trailing whitespaces
Minor cosmetic changes

Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Pascual2005ComputationalEcology; @Petrovskii2012ComputationalEcology].
Thus, they can be seen as experimental systems that allow, unlike to the natural world, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
Due to the usual temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
Here, we are going to focus mainly on exploratory models, but include both pattern- and process-based models.

@@

Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes] and are generally simpler than ecological simulations models [@Hess2020PioLaGPiosphere].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python (NLMpy\; @Etherington2015NLMpyPython), R (NLMR; @Sciaini2018NLMRLandscapetools), Julia (NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia), or stand-alone software (GradientLand; @Cambui2015GradientLandSoftware).

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
<!--JN: software??-->

@@

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behaviors and patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [e.g., @Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling] or connectivity [e.g., @Day2020IndividualbasedModeling].
In order to facilitate future model developments, recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building block [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development history or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

The Digital Twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@deKoning2023DigitalTwins].
Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
<!-- JN: "Digital Twins can" or "Digital Twins will" or "Digital Twins aims to"? -->
Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].
<!--JN: software??-->

>>>>

Simulation models are a powerful tool to study complex, adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
Thus, they can be seen as experimental systems that allow, unlike to the natural world, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
Due to the usual spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
Here, we are going to focus mainly on exploratory models, but include both pattern- and process-based models.

@@

Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes], and are generally simpler than ecological simulations models [@Hess2020PioLaGPiosphere].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels], or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise], or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular], or deforestation [@Soares-Filho2002DinamicaStochastic].
More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].

@@

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [e.g., @Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [e.g., @Day2020IndividualbasedModeling].
In order to facilitate future model developments, recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling], or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

The Digital Twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins], and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
Thus, Digital Twins aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].

### 2024-09-19 Sections/06-genetics.qmd
Minor cosmetic changes

Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], or *landgenreport* [@Gruber2015LandgenreportNew], besides others.
Furthermore, series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].

>>>>

Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], or *landgenreport* [@Gruber2015LandgenreportNew], among others.
Furthermore, series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].

### 2024-09-19 Sections/07-spatial-planning.qmd
Minor cosmetic changes

Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
In the context of landscape-ecology, it provides a pattern-process link<!-- (see section @sec-spatialpattern) -->, and outcomes related to decision making by distinguishing between structure, function and scales [@Hersperger2021LandscapeEcological].
As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology, for example through using simulation model outputs <!-- (see section @sec-simulation) --> as input features  <!-- (see also @daigle2020operationalizing) -->, there have been several recent advances in considering landscape patterns directly in spatial planning.

Generally, spatial patterns influence planning outcomes because of their compactness, contiguity and connectedness.
Furthermore, aggregation can be important because of ecological or practical reasons, for example by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
Novel computational approaches make use of graph theory to identify valuable landscape areas in terms of compact core habitat and reducing boundary exposure [@Weerasena2023SequentialApproach].
Similar, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
In particular, recently the direct consideration of connectivity <!-- (see also section @sec-connectivity) --> has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].

@@

A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathway.
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatial-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
However, this comes with the drawback of reduced interpretability.

<!-- ## Open-source software in spatial conservation planning -->

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
Most software are based on linear or constrained programming approaches requiring a solver to create any outputs.
Unfortunately there is still a gap between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing].

The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows to flexible create planning problems in multiple configurations.
Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
The *restoptr* R package uses constrained programming to identify restoration areas that would most effectively contribute to improved landscape connectivity [@Justeau-Allaire2023RestoptrPackage].
Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
The *CAPTAIN* Python software suite implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
However, many spatial planning software still requires substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.

>>>>

Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
In the context of landscape-ecology, it provides a pattern-process link, and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology.
For example, there have been several recent advances in considering landscape patterns directly in spatial planning, or through using simulation model outputs as input features [e.g., @Daigle2020OperationalizingEcological].

Generally, spatial patterns influence planning outcomes because of their compactness, contiguity and connectedness.
Furthermore, aggregation can be important because of ecological or practical reasons, for example by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
Novel computational approaches make use of graph theory to identify valuable landscape areas in terms of compact core habitat and reducing boundary exposure [@Weerasena2023SequentialApproach].
Similar, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
In particular, recently the direct consideration of connectivity has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].

@@

A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
However, this comes with the drawback of reduced interpretability.

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
Unfortunately there is still a gap between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing].

The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows to flexible create planning problems in multiple configurations.
Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
The *restoptr* R package uses constrained programming to identify restoration areas that would most effectively contribute to improved landscape connectivity [@Justeau-Allaire2023RestoptrPackage].
Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
The *CAPTAIN* Python software library implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
However, many spatial planning software still requires substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.
